Two subtle problems with over-representation analysis

### 2024-07-18 manuscript/manuscript.Rmd
ms work

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
+selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

@@

![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
(A) The number of significant pathways identified in downsampled data that are consistent
with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
(B) The number of significant pathways identified in downsampled data that are not
consistent the full dataset.
(C) The proportion of pathways in downsampled data that are inconsistent with the full
dataset.
All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")

@@

So end-users can explore the effect of these two problems on their work, we have developed
an R/Shiny web application.
Users can upload their foreground and background lists, select their preferred gene set
library and compare results from default and patched analyses.
The resource is currently available at https://oratool.ziemann-lab.net/ and available as a
docker container (https://hub.docker.com/repository/docker/mziemann/background_app/general).

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed with
downsampled real cancer RNA-seq data means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results.
ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that if the data can be scored/ranked, methods like FCS are recommended over ORA.
ORA is better suited to other cases where genes are classified into binary groups.
For example when investigating the over-represented categories in gene expression modules
[@Langfelder2008-hu], or with variant-harbouring genes associated with a disease [@White2019-ps].

@@

Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io),
using the accession numbers mentioned in the Methods section.
The code repository including template Dockerfile and R Markdown script are available on
GitHub (https://github.com/markziemann/background).
The Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/background).
The code repository and Docker image have been uploaded to Zenodo for long-term archiving
(https://zenodo.org/record/TODOXXX).

>>>>

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 0.6, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
log fold change in the case samples of 0.5 for upregulated genes and -0.5 for downregulated
genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to clusterProfiler which suffers from both problems,
clusterProfiler with the background problem fix, clusterProfiler with the FDR problem fix,
clusterProfiler with the fixes to background and FDR problems, an ORA function called `fora`
from `fgsea` bioconductor package v1.30.0 that does not suffer these problems, and `fgsea` a
fast implementation of the preranked gene set enrichment analysis (GSEA) algorithm
[@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest
p-values were used for ORA, like a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.
All analyses were undertaken with R v4.4.0 in a Docker container.

### Comparing ORA and FCS methods by downsampling real RNA-seq data

To understand the sensitivity and false positive rate of ORA and FCS methods, we employed a downsampling
approach to a previously published dataset derived from 37 lung cancer patients, having tumor and normal
tissues for each patient [@Cho2022-ri].
These RNA-seq based gene expression counts were downloaded from NCBI GEO under accession number GSE158420.
After reading them into R, we noticed some gene names converted to dates [@Abeysooriya2021-qt], so we
used the HGNChelper package to fix them [@Oh2022-gj].
DESeq2 was then used for differential expression analysis correcting for patient-of-origin to identify genes
differentially expressed in tumor compared to normal tissue.
FCS was conducted using fgsea using the DESeq2 test statistic to score each gene.
ORA was conducted using fora, using genes with differential expression FDR<0.05.
Separate tests were conducted for up- and down-regulated genes.
For the ORA background, genes with a mean of 10 or more reads per sample on average across the comparison were
included.
Reactome pathways were used.
For downsampling, sample size was varied between 2 and 30 patients, selected pseudorandomly with a set seed,
followed by parallel analysis with fgsea and fora pipelines, using a significance filter of FDR<0.05.
We then calculated the number of pathways identified in downsampled data that were consistent with the full
dataset.
The number of consistent pathways gives an indication of sensitivity of each method.
Then, we calculated the number of pathways identified in the downsampled data that were inconsistent with the
full dataset.
The proportion of significant pathways classified as inconsistent provides an estimate of the real false
discovery rate.
This was repeated 50 times at each sample size, and the median values were reported.

@@

![**Figure 4. Downsampling confirms higher sensitivity for FCS over ORA.**
(A) The number of significant pathways identified in downsampled data that are consistent
with findings in the full dataset, using FCS (fgsea) and ORA (fora) methods.
The blue and red horizontal lines indicate the number of significant pathways identified
in the full set of 27 patients using fgsea and fora respectively.
(B) The number of significant pathways identified in downsampled data that are not
consistent the full dataset.
(C) The proportion of pathways in downsampled data that are inconsistent with the full
dataset.
All values are the median of 50 simulations.](../analysis/fig4.png "Figure 4A shows that fgsea is more sensitive than fora. Figure 4B shows fgsea has higher absule number of false positives, but Figure 4C shows that as a proportion of all findings fgsea has lower false positive rate than fora.")

@@

So end-users can explore the effect of these two problems on their work, we have developed
an R/Shiny web application (see "Availability of data and materials" below).
Users can upload their foreground and background lists, select their preferred gene set
library and compare results from default and patched analyses.

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there
are some bigger problems noted.
Firstly, all variations of ORA were less accurate (based on F1 score) as compared to FCS in this
simulation work, confirming a previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
The higher recall of FCS demonstrated in simulated differential expression profiles and confirmed
with downsampled real cancer RNA-seq data means that researchers will have more power to identify
pathways relevant to the biological processes being studied.
Secondly, many popular ORA tools listed in Table 1 do not provide enrichment scores in their
default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results.
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that if the data can be scored/ranked, methods like FCS are strongly recommended over ORA.
ORA is better suited to other cases where genes are classified into binary groups.
For example when investigating the over-represented categories in gene expression modules
[@Langfelder2008-hu], or with variant-harbouring genes associated with a disease [@White2019-ps].

@@

Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io) and
NCBI GEO using the accession numbers mentioned in the Methods section.

Reproduction of the results here can be achieved using the GitHub code repository GitHub
(https://github.com/markziemann/background) and Docker image
(https://hub.docker.com/r/mziemann/background), which have both been deposited to Zenodo
for long-term preservation (https://zenodo.org/record/TODOXXX).

The R/Shiny tool for comparing ORA with and without the two problems is currently
online at https://oratool.ziemann-lab.net/ and available as a
downloadable docker container for local execution
(https://hub.docker.com/repository/docker/mziemann/background_app/general).
